cardinalR: Generating interesting high-dimensional data structures

### 2025-04-30 paper-cardinalR.Rmd
Add figure references for the application

We generated a dataset of five clusters, positioned with equal inter-cluster distances in $4\text{-}D$ space. The five clusters have different geometric structures and each contain different number of points. Specifically, the helical spiral cluster includes $2250$ points and was generated with a scale parameter of $0.4$. The hemisphere cluster consists of $1500$ points with a scale parameter of $0.35$. The uniform cube-shaped cluster contains $750$ points and uses a scale parameter of $0.3$. The blunted cone cluster includes $1250$ points, generated with a scale parameter of $1$. Finally, the Gaussian-shaped cluster contains $1750$ points and was generated with a scale parameter of $0.3$.

@@

@@

UMAP, PHATE, TriMAP, and PaCMAP effectively separate the five clusters and show the preservation of the global structure. However, PHATE reveals three non-linear clusters, even though two of them do not show non-linearity. UMAP, TriMAP, and PaCMAP successfully maintain the local structures of the data. In contrast, tSNE divides the non-linear cluster into sub-clusters. Also, tSNE fails to preserve the distances between the clusters. PCA, on the other hand, preserves the local structures of the clusters, but some clusters are incorrectly merged that should remain distinct.

>>>>

The following code generates a dataset of five clusters, positioned with equal inter-cluster distances in $4\text{-}D$ space (Figure \@ref(fig:fig-highd-proj)).

@@

The five clusters have different geometric structures and each contain different number of points. Specifically, the helical spiral cluster includes $2250$ points and was generated with a scale parameter of $0.4$. The hemisphere cluster consists of $1500$ points with a scale parameter of $0.35$. The uniform cube-shaped cluster contains $750$ points and uses a scale parameter of $0.3$. The blunted cone cluster includes $1250$ points, generated with a scale parameter of $1$. Finally, the Gaussian-shaped cluster contains $1750$ points and was generated with a scale parameter of $0.3$.

@@

UMAP, PHATE, TriMAP, and PaCMAP effectively separate the five clusters and show the preservation of the global structure (Figure \@ref(fig:fig-nldr-layouts)). However, PHATE reveals three non-linear clusters, even though two of them do not show non-linearity. UMAP, TriMAP, and PaCMAP successfully maintain the local structures of the data. In contrast, tSNE divides the non-linear cluster into sub-clusters. Also, tSNE fails to preserve the distances between the clusters. PCA, on the other hand, preserves the local structures of the clusters, but some clusters are incorrectly merged that should remain distinct.
